---
title: "IBNR Estimation — Four Actuarial Methods"
output:
  html_document:
    toc: true
    toc_float: true
    theme: flatly
    highlight: tango
date: "`r Sys.Date()`"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 10, fig.height = 5)
```

This notebook estimates **Incurred But Not Reported (IBNR)** reserves using four standard actuarial methods.

**Methods:**

1. Chain Ladder (Development Method)
2. Average Development Method
3. Bornhuetter-Ferguson Method
4. Cape Cod Method

---

## 1. Load Data

```{r load-data}
library(ggplot2)
library(dplyr)
library(tidyr)
library(knitr)
library(scales)

# Load triangle
tri_raw <- read.csv("../data/paid_loss_triangle.csv", check.names = FALSE)
rownames(tri_raw) <- tri_raw$accident_year
tri_raw$accident_year <- NULL
colnames(tri_raw) <- as.integer(gsub("dev_", "", colnames(tri_raw)))

# Load premium
premium_df <- read.csv("../data/earned_premium.csv")
AYs <- as.integer(rownames(tri_raw))
devs <- as.integer(colnames(tri_raw))
n <- length(AYs)

# Matrix form
triangle <- as.matrix(tri_raw)
premium <- setNames(premium_df$earned_premium, premium_df$accident_year)
premium <- premium[as.character(AYs)]

kable(tri_raw, format.args = list(big.mark = ","), caption = "Paid Loss Triangle ($000s)")
```

---

## 2. Age-to-Age Development Factors

```{r ata-factors}
# Volume-weighted ATA factors
ata_vw <- numeric(n - 1)
ata_sa <- numeric(n - 1)

for (j in seq_len(n - 1)) {
  col_curr <- triangle[, j]
  col_next <- triangle[, j + 1]
  mask <- !is.na(col_curr) & !is.na(col_next)
  if (sum(mask) > 0) {
    ata_vw[j] <- sum(col_next[mask]) / sum(col_curr[mask])
    ata_sa[j] <- mean(col_next[mask] / col_curr[mask])
  }
}

# Individual link ratios
lr_mat <- matrix(NA, nrow = n, ncol = n - 1,
                 dimnames = list(AYs, paste0(devs[-n], "-", devs[-1])))
for (i in seq_len(n)) {
  for (j in seq_len(n - 1)) {
    if (!is.na(triangle[i, j]) && !is.na(triangle[i, j + 1])) {
      lr_mat[i, j] <- triangle[i, j + 1] / triangle[i, j]
    }
  }
}

lr_display <- rbind(lr_mat, "Vol. Wtd. Avg" = ata_vw, "Simple Avg" = ata_sa)
kable(round(lr_display, 4), caption = "Link Ratios by Accident Year")
```

```{r cdfs}
# Cumulative development factors from each period to ultimate
TAIL <- 1.0
cdfs_from_ata <- function(ata, tail = 1.0) {
  factors <- c(ata, tail)
  cdfs <- numeric(length(factors))
  cdfs[length(cdfs)] <- tail
  for (i in (length(factors) - 1):1) {
    cdfs[i] <- factors[i] * cdfs[i + 1]
  }
  cdfs
}

cdfs_vw <- cdfs_from_ata(ata_vw, TAIL)
cdfs_sa <- cdfs_from_ata(ata_sa, TAIL)

cdf_df <- data.frame(
  Development = paste0(devs, "m"),
  `CDF (Vol. Wtd.)` = round(cdfs_vw, 4),
  `CDF (Simple Avg)` = round(cdfs_sa, 4),
  `% Developed` = paste0(round(100 / cdfs_vw, 1), "%"),
  check.names = FALSE
)
kable(cdf_df, caption = "Cumulative Development Factors")
```

---

## 3. Method 1: Chain Ladder

Projects the latest diagonal using volume-weighted CDF:

$$\hat{U}_i = C_{i,d_i} \times CDF_{d_i}$$

```{r chain-ladder}
# Find latest diagonal
latest <- numeric(n)
latest_col <- integer(n)
for (i in seq_len(n)) {
  known <- which(!is.na(triangle[i, ]))
  if (length(known) > 0) {
    j <- max(known)
    latest[i] <- triangle[i, j]
    latest_col[i] <- j
  }
}

# Chain Ladder
ult_cl <- latest * cdfs_vw[latest_col]
ibnr_cl <- ult_cl - latest

cl_df <- data.frame(
  `Accident Year` = AYs,
  `Latest Paid` = latest,
  `CDF` = round(cdfs_vw[latest_col], 4),
  `Ultimate` = round(ult_cl),
  `IBNR` = round(ibnr_cl),
  check.names = FALSE
)

cl_total <- data.frame(
  `Accident Year` = "TOTAL", `Latest Paid` = sum(latest), `CDF` = NA,
  `Ultimate` = round(sum(ult_cl)), `IBNR` = round(sum(ibnr_cl)),
  check.names = FALSE
)
kable(rbind(cl_df, cl_total), format.args = list(big.mark = ","),
      caption = "Chain Ladder Results ($000s)")
```

---

## 4. Method 2: Average Development

Same as Chain Ladder but using simple average link ratios:

```{r avg-development}
ult_ad <- latest * cdfs_sa[latest_col]
ibnr_ad <- ult_ad - latest

ad_df <- data.frame(
  `Accident Year` = AYs,
  `Latest Paid` = latest,
  `CDF (Simple Avg)` = round(cdfs_sa[latest_col], 4),
  `Ultimate` = round(ult_ad),
  `IBNR` = round(ibnr_ad),
  check.names = FALSE
)
ad_total <- data.frame(
  `Accident Year` = "TOTAL", `Latest Paid` = sum(latest), `CDF (Simple Avg)` = NA,
  `Ultimate` = round(sum(ult_ad)), `IBNR` = round(sum(ibnr_ad)),
  check.names = FALSE
)
kable(rbind(ad_df, ad_total), format.args = list(big.mark = ","),
      caption = "Average Development Results ($000s)")
```

---

## 5. Method 3: Bornhuetter-Ferguson

Blends Chain Ladder with an *a priori* expected loss ratio:

$$\hat{U}_i^{BF} = C_{i,d_i} + \left(1 - \frac{1}{CDF_{d_i}}\right) \times ELR \times P_i$$

```{r bf}
# ELR from Chain Ladder ultimates
ELR <- sum(ult_cl) / sum(premium)
cat(sprintf("A priori ELR (from Chain Ladder): %.1f%%\n", ELR * 100))

unrep_pct <- 1 - 1 / cdfs_vw[latest_col]
exp_losses <- ELR * premium
ibnr_bf <- unrep_pct * exp_losses
ult_bf <- latest + ibnr_bf

bf_df <- data.frame(
  `Accident Year` = AYs,
  `Latest Paid` = latest,
  `Unreported %` = paste0(round(unrep_pct * 100, 1), "%"),
  `Expected Losses` = round(exp_losses),
  `IBNR` = round(ibnr_bf),
  `Ultimate` = round(ult_bf),
  check.names = FALSE
)
bf_total <- data.frame(
  `Accident Year` = "TOTAL", `Latest Paid` = sum(latest), `Unreported %` = NA,
  `Expected Losses` = round(sum(exp_losses)), `IBNR` = round(sum(ibnr_bf)),
  `Ultimate` = round(sum(ult_bf)), check.names = FALSE
)
kable(rbind(bf_df, bf_total), format.args = list(big.mark = ","),
      caption = "Bornhuetter-Ferguson Results ($000s)")
```

---

## 6. Method 4: Cape Cod

ELR estimated from the data via *used-up premium*:

$$ELR^{CC} = \frac{\sum_i C_{i,d_i}}{\sum_i P_i / CDF_{d_i}}$$

```{r cape-cod}
used_up_prem <- premium / cdfs_vw[latest_col]
ELR_cc <- sum(latest) / sum(used_up_prem)
cat(sprintf("Cape Cod ELR (from data): %.1f%%\n", ELR_cc * 100))

unrep_pct_cc <- 1 - 1 / cdfs_vw[latest_col]
ibnr_cc <- unrep_pct_cc * ELR_cc * premium
ult_cc <- latest + ibnr_cc

cc_df <- data.frame(
  `Accident Year` = AYs,
  `Latest Paid` = latest,
  `Used-Up Premium` = round(used_up_prem),
  `Unreported %` = paste0(round(unrep_pct_cc * 100, 1), "%"),
  `IBNR` = round(ibnr_cc),
  `Ultimate` = round(ult_cc),
  check.names = FALSE
)
cc_total <- data.frame(
  `Accident Year` = "TOTAL", `Latest Paid` = sum(latest),
  `Used-Up Premium` = round(sum(used_up_prem)), `Unreported %` = NA,
  `IBNR` = round(sum(ibnr_cc)), `Ultimate` = round(sum(ult_cc)),
  check.names = FALSE
)
kable(rbind(cc_df, cc_total), format.args = list(big.mark = ","),
      caption = "Cape Cod Results ($000s)")
```

---

## 7. Results Comparison

```{r comparison-table}
comp <- data.frame(
  `Accident Year` = AYs,
  `Latest Paid` = latest,
  `Chain Ladder` = round(ibnr_cl),
  `Avg Development` = round(ibnr_ad),
  `Bornhuetter-Ferguson` = round(ibnr_bf),
  `Cape Cod` = round(ibnr_cc),
  check.names = FALSE
)
comp_total <- data.frame(
  `Accident Year` = "TOTAL", `Latest Paid` = sum(latest),
  `Chain Ladder` = round(sum(ibnr_cl)), `Avg Development` = round(sum(ibnr_ad)),
  `Bornhuetter-Ferguson` = round(sum(ibnr_bf)), `Cape Cod` = round(sum(ibnr_cc)),
  check.names = FALSE
)
kable(rbind(comp, comp_total), format.args = list(big.mark = ","),
      caption = "IBNR by Method ($000s)")
```

```{r comparison-chart}
# Reshape for ggplot
ibnr_long <- data.frame(
  `Accident Year` = rep(as.character(AYs), 4),
  Method = rep(c("Chain Ladder", "Avg Development", "Bornhuetter-Ferguson", "Cape Cod"),
               each = n),
  IBNR = c(ibnr_cl, ibnr_ad, ibnr_bf, ibnr_cc),
  check.names = FALSE
)
ibnr_long$Method <- factor(ibnr_long$Method,
                           levels = c("Chain Ladder", "Avg Development",
                                      "Bornhuetter-Ferguson", "Cape Cod"))

ggplot(ibnr_long, aes(x = `Accident Year`, y = IBNR, fill = Method)) +
  geom_bar(stat = "identity", position = "dodge", alpha = 0.85) +
  scale_y_continuous(labels = comma) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "IBNR by Accident Year and Method",
       y = "IBNR ($000s)", x = "Accident Year") +
  theme_minimal(base_size = 12) +
  theme(legend.position = "bottom")
```

```{r totals-chart}
totals_df <- data.frame(
  Method = c("Chain Ladder", "Avg Development", "Bornhuetter-Ferguson", "Cape Cod"),
  IBNR = c(sum(ibnr_cl), sum(ibnr_ad), sum(ibnr_bf), sum(ibnr_cc))
)
totals_df$Method <- factor(totals_df$Method, levels = totals_df$Method)

ggplot(totals_df, aes(x = Method, y = IBNR, fill = Method)) +
  geom_bar(stat = "identity", alpha = 0.85, show.legend = FALSE) +
  geom_text(aes(label = comma(round(IBNR))), vjust = -0.5, fontface = "bold") +
  scale_y_continuous(labels = comma, expand = expansion(mult = c(0, 0.15))) +
  scale_fill_brewer(palette = "Set1") +
  labs(title = "Total IBNR by Method", y = "Total IBNR ($000s)", x = NULL) +
  theme_minimal(base_size = 12)
```

---

## 8. Summary

```{r summary}
sum_df <- data.frame(
  Method = c("Chain Ladder", "Avg Development", "Bornhuetter-Ferguson", "Cape Cod"),
  `Total IBNR` = round(c(sum(ibnr_cl), sum(ibnr_ad), sum(ibnr_bf), sum(ibnr_cc))),
  `Total Ultimate` = round(c(sum(ult_cl), sum(ult_ad), sum(ult_bf), sum(ult_cc))),
  `Total Premium` = rep(sum(premium), 4),
  check.names = FALSE
)
sum_df$`Ultimate LR` <- paste0(round(sum_df$`Total Ultimate` / sum_df$`Total Premium` * 100, 1), "%")
sum_df$`IBNR / Latest` <- paste0(round(sum_df$`Total IBNR` / sum(latest) * 100, 1), "%")

kable(sum_df[, c("Method", "Total IBNR", "Total Ultimate", "Ultimate LR", "IBNR / Latest")],
      format.args = list(big.mark = ","), caption = "Summary")
```

---

## Method Reference

| Method | Strengths | Weaknesses |
|---|---|---|
| **Chain Ladder** | Simple, data-driven, widely used | Amplifies distortions in early diagonals |
| **Average Development** | Equal weight per year; smooths volume bias | May underweight credible large-volume years |
| **Bornhuetter-Ferguson** | Stabilises immature years; blends prior + data | Sensitive to ELR choice |
| **Cape Cod** | ELR derived from data; no external assumption | Volatile if experience is thin |
